---
title: "Lecture 7.2: Basic linear regression"
output: pdf_document
fontsize: 12pt 
geometry: margin=1in
---

```{r setup, include=FALSE}
library(knitr)
library(extrafont)
loadfonts()
setwd("~/Dropbox/BayesClass/2017 Class/Lecture 7")
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  fig.align = "center",
  fig.height = 3,
  fig.width = 4
  )
```
\emph{* This lecture is based on chapter 3 of Statistical Rethinking by Richard McElreath.}

As before, we need to load some packages and set some options prior to running any models:

```{r stanPackages, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(rstan)
library(shinystan)
library(car)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
source("../utilityFunctions.R")
```

\section{Data story}
\emph{Dictyota menstrualis} is a brown seaweed that produces > 250 dichyol terpenes as chemical deterrents against marine herbivores and biofouling microbes. 

  * When herbivore pressure is high, having higher terpene concentrations should lead to higher lifetime biomass. At least that's my story! 

I have saved this data as a `.csv` file called `algae.csv`. Also, below is a function, `seaweedSim` with the code to simulate your own data if you want to play around with it.
```{r, eval=FALSE}
# You give it the total number of desired observations, the intercept, 
# the slope, and the standard deviation and it makes a dataframe for you. 
# Play with it, changing parameters as you wish to see how the model
# differs. Also, in the script, terpenes is the x variable. I didn't
# include arguments to change that, but it would be easy by changing
# the mean=50 & sd=3 to whatever you want.

seaweedSim <- function(nObs, alpha, beta, sigma) {
  terpenes <- round(rnorm(nObs, mean=50, sd=3), digits=2) 
  error <- rnorm(nObs, mean=0, sd=sigma)
  biomass <- alpha + beta * terpenes + error
  out <- data.frame(terpenes, biomass)
  out <- out[order(terpenes),]
  return(out)
}

set.seed(20)
algae <- seaweedSim(nObs=50, alpha=0, beta=3, sigma=12)
write.csv(algae, file = "algae.csv", row.names = FALSE)
```

```{r}
algae <- read.csv("algae.csv")
head(algae)
```

\section{Modeling the dependent variable with a Gaussian distribution}

We will begin with a single measurement variable, `biomass`, to model as a normal distribution. There are two parameters describing the distribution's shape:

  1. $\mu$: the mean describing the central location
  2. $\sigma$: the standard deviation describing the spread

Bayes and MCMC will allow us to explore a number of the most plausible distributions, each with their own $\mu$ and $\sigma$ and rank them by their posterior plausability. 

To define our model for biomass as normally distributed with mean $\mu$ and standard deviation $\sigma$, we need to define a prior $\mathrm{Pr}(\mu, \sigma)$---the \emph{joint prior probability} for the parameters. 
\begin{itemize}
  \item For many purposes, priors are specified independently for each parameter (as we have done previously). Thus we assume $\mathrm{Pr}(\mu, \sigma) = \mathrm{Pr}(\mu) \mathrm{Pr}(\sigma)$.
\end{itemize}

\begin{align}
  BM_i    &\sim \mathrm{Normal}(\mu, \sigma)  \nonumber  \\ 
  \mu     &\sim \mathrm{Normal}(150, 30)                 \\
  \sigma  &\sim \mathrm{Cauchy^+}(0, 10)      \nonumber
\end{align}

I have set the priors as follows: 

  * The prior for $\mu$ is weakly informative and centered on the mean of `biomass` with 95% probability the average is between $150 \pm 60$. 
    * Later we will play with more restrictive \emph{regularizing priors}
    
  * The prior for $\sigma$ must be positive. A uniform distibution distribution for this usually doesn't sample well from $\mathrm{U}(0, \infty)$, and setting upper bounds on uniforms can cause issues. 
  
  * instead we will use a \emph{half-Cauchy} distribution. This is equivalent to a folded t-distribution with $df=1$. 
    * It centers most of the probability mass around zero and therefore credible values, but has fat tails for extremes. You can play with the Cauchy with the `dcauchy` function. 

```{r, fig.height=2.75, fig.width=4, echo=-1}
par(mar=c(3,3,0.1,0.5))
curve(dnorm(x, 0, 5), from=0, to=20, las=1)
curve(dcauchy(x, 0, 5), add=TRUE, col="blue")
curve(dnorm(x, 0, 10), add=TRUE, col="red")
curve(dcauchy(x, 0, 10), add=TRUE, col="cornflowerblue")

mtext(text = expression(bold(sigma)), side=1, line = 2)
text(13.5, 0.075, "Normal (0,5)", font=1,cex=1, col="black", adj=c(0, 0.5))
text(13.5, 0.0675, "Cauchy (0,5)", font=1,cex=1, col="blue", adj=c(0, 0.5))
text(13.5, 0.06, "Normal (0,10)", font=1,cex=1, col="red", adj=c(0, 0.5))
text(13.5, 0.0525, "Cauchy (0,10)", font=1,cex=1, col="cornflowerblue", 
  adj=c(0, 0.5))
```    
    
We set up our model (`modMean.stan`) similarly to how we set up the simple binomial models previously, except that our data are now part of a vector. 

```{r, eval=FALSE}
data {
  int<lower=0> nObs;          // No. obs.
  vector<lower=0>[nObs] BM;   // biomass observations
  real<lower=0> muMean;       // mean of prior mu
  real<lower=0> muSD;         // SD of prior mu
  real<lower=0> sigmaSD;      // scale for sigma
}  

parameters {
  real mu;
  real<lower=0> sigma;
}

model {
  mu ~ normal(muMean, muSD);
  sigma ~ cauchy(0, sigmaSD);
  
  BM ~ normal(mu, sigma);
}

```
```{r engine = 'cat', engine.opts = list(file = "modMean.stan", lang = "stan"), echo=FALSE}
data {
  int<lower=0> nObs;          // No. obs.
  vector<lower=0>[nObs] BM;   // biomass observations
  real<lower=0> muMean;       // mean of prior mu
  real<lower=0> muSD;         // SD of prior mu
  real<lower=0> sigmaSD;      // scale for sigma
}  

parameters {
  real mu;
  real<lower=0> sigma;
}

model {
  mu ~ normal(muMean, muSD);
  sigma ~ cauchy(0, sigmaSD);
  
  BM ~ normal(mu, sigma);
}

```

Lets set up the data and look at the simplest model first:
```{r, message=FALSE, warning=FALSE, cache=TRUE, verbose=FALSE}

dat <- list(nObs=dim(algae)[1], BM=algae$biomass, muMean=150, 
            muSD=30, sigmaSD=10)

intMod <- stan(file="modMean.stan", data=dat, iter=2000, chains=4, seed=3)

parMod <- as.data.frame(intMod, pars=c("mu", "sigma"))
```

```{r}
print(intMod, pars=c("mu", "sigma"), digits.summary=2)
```



We can plot the marginal densities and 95% HDI's of $\mu$ &  $\sigma$: 


```{r, fig.height=3, fig.width=7, echo=c(-1,-2)}
par(mar=c(3,3,0.15,0.5))
par(mfrow=c(1,2))

plotInterval(parMod$mu, HDI=TRUE, credMass=0.95, xlims=c(140, 160),
              col="blue", yOffset=0.01)
mtext(expression(paste(bold(mu))), side=1, line=2, cex=1.2)

plotInterval(parMod$sigma, HDI=TRUE, credMass=0.95, xlims=c(10, 25), 
             col="blue", yOffset=0.01)
mtext(expression(paste(bold(sigma))), side=1, line=2, cex=1.2)
``` 
  
Or plot the joint posterior density $\mathrm{Pr}(\mu, \sigma)$:

```{r, fig.height=2.8, fig.width=4, echo=-1}
par(mar=c(3,3,0.1,0.5))
col <- "#50505010"
plot(parMod, pch=16, col=col, las=1, ylim=c(10,25), 
      xlim=c(140,160), bty="l")
dataEllipse(as.matrix(parMod),level=c(0.25,0.5,0.95), add=TRUE, labels=FALSE,
            plot.points=FALSE, center.pch=FALSE, col=c(col,"#006DCC"))
mtext(text = expression(paste(sigma)), side=2, line=2.2, cex=1.2, las=1)
mtext(text = expression(paste(mu)), side=1, line=2, cex=1.2)
```  

As before, if we want to estimate the biomass for the \emph{Dictyota} population, we need to consider both the posterior mean and standard deviations.

```{r, fig.height=3, fig.width=4, echo=-1}
par(mar=c(3,3.2,0.1,0.5))
# plot empty plot
plot(0:1,0:1, type="n", xlim=c(100, 200), ylim=c(0,0.035), las=1, bty="l") 
mtext(text = "Estimated biomass", side=1, line = 2, cex=1)

# Overlay posterior biomass densities
for (n in 1:nrow(parMod)) {
  curve(dnorm(x, parMod[n,1], parMod[n,2]), add=TRUE, col="#50505010")
}

# Overlay median posterior probability density
medBM <- apply(parMod,2,median)
curve(dnorm(x,medBM[1], medBM[2]), add=TRUE, col="cornflowerblue", lwd=3)
```

For this intercept only model, $\mu$ and $\sigma$ are relatively uncorrelated. 
```{r, digits=2}
cor(parMod)
```


This can change though once we add a predictor. 

\section{Linear modeling}

Modeling the average biomass in the population is interesting, but usually we are interested in how biomass changes as a function of some other variable(s), such as terpene concentration. 

The simplest way to do this is with a linear model. In linear models, the predictor variable has a constant additive relationship with the mean of the outcome. 

By adding terpenes ($x_i$), the model can be reformulated as:

\begin{align}
  BM_i    &\sim \mathrm{Normal}(\mu_i, \sigma)  \nonumber  \\ 
  \mu_i     &= \alpha + \beta x_i        \nonumber  \\         
  \alpha  &\sim \mathrm{Normal}(150, 100)                 \\
  \beta   &\sim \mathrm{Normal}(0, 10)        \nonumber   \\
  \sigma  &\sim \mathrm{Cauchy^+}(0, 10)      \nonumber
\end{align}

Now $\mu$ is no longer a paremter to be estimated but instead is a deterministic function of two parameters and a variable, where the two parameters are:
  
  * $\alpha$: the expected biomass when there are zero terpenes
  * $\beta$: The expected change in biomass when terpene concentrations are increased by one unit.
  
We have specified a wide prior for $\alpha$ because it often can take a wide range of values. 

The prior for beta specifies that we have no \emph{a priori} expectation that $\beta$ should be positive or negative, while accepting a relatively wide range of values. 

In Stan, coding this should be easy. Try it and call the file `modLM.stan`.

```{r, eval=FALSE}
data {
  int<lower=0> nObs;          // No. obs.
  vector<lower=0>[nObs] BM;   // biomass observations
  vector<lower=0>[nObs] terpenes;
  real<lower=0> aMean;       // mean of prior alpha
  real<lower=0> aSD;         // SD of prior alpha
  real<lower=0> bSD;          // SD of prior beta
  real<lower=0> sigmaSD;      // scale for sigma
}

parameters {
  real alpha;
  real beta;
  real<lower=0> sigma;
}

transformed parameters {
    // can be useful for plotting purposes
  vector[nObs] mu;
  mu = alpha + beta*terpenes;
}

model {
  alpha ~ normal(aMean, aSD);
  beta ~  normal(0, bSD);
  sigma ~ cauchy(0, sigmaSD);

  BM ~ normal(mu, sigma);
}

```
```{r engine = 'cat', engine.opts = list(file = "modLM.stan", lang = "stan"), echo=FALSE}
data {
  int<lower=0> nObs;          // No. obs.
  vector<lower=0>[nObs] BM;   // biomass observations
  vector<lower=0>[nObs] terpenes;
  real<lower=0> aMean;       // mean of prior alpha
  real<lower=0> aSD;         // SD of prior alpha
  real<lower=0> bSD;          // SD of prior beta
  real<lower=0> sigmaSD;      // scale for sigma
}

parameters {
  real alpha;
  real beta;
  real<lower=0> sigma;
}

transformed parameters {
  // can be useful for plotting purposes
  vector[nObs] mu;
  mu = alpha + beta*terpenes;
}

model {
  alpha ~ normal(aMean, aSD);
  beta ~  normal(0, bSD);
  sigma ~ cauchy(0, sigmaSD);

  BM ~ normal(mu, sigma);
}

```

```{r, message=FALSE, warning=FALSE, cache=TRUE, verbose=FALSE}

datLM <- list(nObs=dim(algae)[1], BM=algae$biomass, terpenes=algae$terpenes,
              aMean=150, aSD=100, bSD=10, sigmaSD=10)

modLM <- stan(file="modLM.stan", data=datLM, iter=2000, chains=4, seed=3)
```


